---
title: "Stat107 Final Project"
subtitle: "How do harmful air pollutant concentrations vary across different months within a year?"
author: "Crystal Kuong, Gracelynne Mohad, Sumeet Sammi, Zoe Shum"
date: "2024-12-13"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-line-numbers: true
    page-layout: full
  pdf:
    fontsize: 10pt
    code-line-numbers: true
---

# Introduction

Air quality is a critical component in environmental health, directly influencing both the ecosystem and human well-being. It is determined by the presence of various pollutants including, but not limited to nitrogen dioxide and carbon monoxide. Poor air quality is linked to serious health issues, including respiratory diseases, cardiovascular problems, and reduced life expectancy. Moreover, air pollution contributes to climate change and harming wildlife. Monitoring and understanding air quality prompts initiatives centered on reducing the concentration of air pollutants, thereby, creating a cleaner environment.

The specific pollutants we will explore in this study include:

1.  **Carbon Monoxide** ($CO$): A colorless, odorless gas that is produced by the incomplete combustion of fossil fuels, which contributes to air pollution and can be harmful to humans as it blocks the successful delivery of oxygen to vital areas.  

2.  **Non-Methane Hydrocarbon** ($NMHC$)**:** Organic compounds excluding methane that contribute to the formation of ground level ozone and secondary air pollutants through photochemical reactions. 

3.  **Nitrogen Dioxide** ($NO_2$)**:** A reddish-brown colored gas that is produced by combustion of fossil fuels, contributes to acid rain, smog, and respiratory problems in humans.

4.  **Nitrogen Oxide** ($NOx$)**:** Similarly to $NO_2$, a gas that is produced by combustion of fossil fuels, contributes to acid rain, smog, and ground level ozone.

5.  **Ozone** ($O_3$)**:** A secondary pollutant formed in the atmosphere from precursor emissions like $NOx$ and volatile organic compounds.

## The Dataset

The UCI Air Quality Dataset was collected by the University of California, Irvine and acquired from Kaggle. It measures the concentration of about 5 different pollutants in the air in a given area. Some pollutants that are measured include carbon monoxide, benzene, and nitrogen dioxide. It measures pollutant concentrations every hour from March 2004 to April 2005.

### Packages

```{r}
library(readxl)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(dplyr)
```

```{r}
#| code-fold: true
air <- read_excel("AirQualityUCI.xls")
head(air)
```

The original dataset contains 9,357 observations and 15 variables. Variable columns 3-12 will be renamed as followed to enhance clarity.

```{r}
#| code-fold: true
air <- air %>% 
  rename(
    CO = `CO(GT)`,
    S_CO = `PT08.S1(CO)`,
    NMHC = `NMHC(GT)`,
    S_NMHC = `PT08.S2(NMHC)`,
    C6H6 = `C6H6(GT)`,
    NOx = `NOx(GT)`,
    S_NOx = `PT08.S3(NOx)`,
    NO2 = `NO2(GT)`,
    S_NO2 = `PT08.S4(NO2)`,
    S_O3 = `PT08.S5(O3)`
  )
```

| Variable | Description                                                    |
|--------------------|----------------------------------------------------|
| `Date`   | The date of the measurement                                    |
| `Time`   | The time of the measurement.                                   |
| `CO`     | Concentration of carbon monoxide ($CO$) in the air (µg/m³).    |
| `S_CO`   | Sensor measurement for $CO$ concentration.                     |
| `NMHC`   | Concentration of non-methane hydrocarbons ($NMHC$) (µg/m³).    |
| `C6H6`   | Concentration of benzene ($C_6H_6$) in the air (µg/m³).        |
| `S_NMHC` | Sensor measurement for NMHC concentration.                     |
| `NOx`    | Concentration of nitrogen oxide ($NOx$) in the air (µg/m³).    |
| `S_NOx`  | Sensor measurement for $NOx$ concentration.                    |
| `NO2`    | Concentration of nitrogen dioxide ($NO_2$) in the air (µg/m³). |
| `S_NO2`  | Sensor measurement for $NO_2$ concentration.                   |
| `S_O3`   | Sensor measurement for ozone ($O_3$) concentration             |
| `T`      | Temperature (Celsius)                                          |
| `RH`     | Relative humidity (g/m³)                                       |
| `AH`     | Absolute humidity (g/m³)                                       |

### Tidying The Dataset

```{r}
#| code-fold: true
#Summarize dataset (excluding Date and Time columns)
summary(air[, -c(1, 2)])
```

```{r}
#| code-fold: true
# Histogram for S_CO 
hist(air$S_CO,
     main = "Histogram of CO concentrations",
     xlab = "CO concentrations")
```

Based on the summary statistics, an unusual minimum value of -200.00 was observed across all variables. To investigate this anomaly, the variable `S_CO` was analyzed as a test case. A histogram revealed the presence of significant outliers, including values around -200.00. To address these issues and enhance data quality, outliers were identified and removed using the **Interquartile Range (IQR) method**.

```{r}
#| code-fold: true
# Calculate IQR
Q1 <- quantile(air$S_CO, 0.25, na.rm = TRUE)
Q3 <- quantile(air$S_CO, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

# Define outlier bounds
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter rows to remove outliers
air_filtered <- air[air$S_CO >= lower_bound & air$S_CO <= upper_bound, ]
```

The outcome:

-   484 total observations were removed.

-   The distribution of **`S_CO`** shifted closer to normality, although still retaining a slight positive skew.

```{r}
#| code-fold: true
hist(air_filtered$S_CO,
     main = "Histogram of CO concentrations (after filtering)",
     xlab = "CO concentrations")
```

### Reformatting The Dataset

To effectively analyze monthly variations in pollutant concentrations, the dataset will be reorganized by grouping the data by month and year, calculating the average values for each variable within these groups. This restructuring will provide a clearer understanding of seasonal trends and fluctuations.

```{r}
#| code-fold: true
# Reassign original dataset with filtered dataset
air <- air_filtered

# Convert 'Date' column to y-m-d format
air$Date <- ymd(air$Date)

# Create numerical Month and Year columns
air <- air %>%
  mutate(
    Month = month(Date, label = TRUE), # Extracts month as a numeric value (1-12)
    Year = year(Date)    # Extracts year
  )

# Group by Year and Month for summarization
airM <- air %>%
  group_by(Year, Month) %>%
  summarize(across(where(is.numeric), 
                   ~mean(., na.rm = TRUE)),
            .groups = "drop")

head(airM, 5)
```

Upon inspecting the dataset, some non-sensor measurements (e.g., CO, NMHC) revealed unusual negative values, suggesting potential inaccuracies. For this reason, the analysis will prioritize **sensor-based variables**.

```{r}
#| code-fold: true
airM <- air[ ,-c(3,5,6,8,10)] %>%
  group_by(Year, Month) %>%
  summarize(across(where(is.numeric), 
                   ~mean(., na.rm = TRUE)),
            .groups = "drop")

head(airM, 5)
```

# Analysis

```{r fig.width=10, fig.height=6}
#| code-fold: true
# Plot monthly Concentrations
airM %>%
  # Use interaction to combine Month and Year
  ggplot(aes(x = interaction(Month, Year, sep = "-"))) +
  # Add points and lines for different pollutants
  geom_line(aes(y = S_CO, color = "Carbon Monoxide", group = 1)) +
  geom_line(aes(y = S_NMHC, color = "Non-Methane Hydrocarbon", group = 2)) +
  geom_line(aes(y = S_NOx, color = "Nitrogen Oxide", group = 3)) +
  geom_line(aes(y = S_NO2, color = "Nitrogen Dioxide", group = 3)) +
  geom_line(aes(y = S_O3, color = "Ozone", group = 3)) +
  labs(
    title = "Average Concentrations of Pollutants by Month (2004-2005)",
    x = "Month-Year",
    y = "Concentrations in µg/m³"
  ) +
  scale_color_manual(
    values = c("Carbon Monoxide" = "dodgerblue3", 
               "Non-Methane Hydrocarbon" = "brown2", 
               "Nitrogen Oxide" = "palegreen3",
               "Nitrogen Dioxide" = "mediumpurple3",
               "Ozone" = "hotpink2"),
    name = "Pollutant"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels
```

```{r fig.width=10, fig.height=6}
#| code-fold: true
# Pivot data frame by Pollutant 
airM_2 <- airM %>%
  pivot_longer(
    cols = c(3:7), 
    names_to = "Pollutant", 
    values_to = "Concentration"
  )

# Plot Pollutant Concentrations 
airM_2 %>% 
  ggplot(aes(x = Pollutant)) +
  geom_boxplot(aes(y = Concentration, fill = Pollutant), size = 0.5) +
  labs(
    title = "Concentrations of Pollutants",
    x = "Pollutant",
    y = "Concentration (µg/m³)"
  ) +
  scale_fill_manual(
    values = c("dodgerblue3", "brown2", "mediumpurple3", "palegreen3", "hotpink2"), # Specify colors for each pollutant
    labels = c("Carbon Monoxide", "Non-methane Hydrocarbons", "Nitrogen Dioxide", "Nitrogen Oxide", "Ozone")
  )
```

```{r fig.width=10, fig.height=6}
#| code-fold: true
# Plot monthly Temperatures & Relative Humidity 
airM %>%
  pivot_longer(
    cols = c(RH, T), 
    names_to = "Variable", 
    values_to = "Value"
  ) %>%
  ggplot(aes(x = interaction(Month, Year, sep = "-"), y = Value, fill = Variable)) +
  geom_col(position = "dodge") +
  labs(
    title = "Temperature and Relative Humidity by Month",
    x = "Month-Year",
    y = "Values",
    fill = "Variable"
  ) +
  scale_fill_manual(
    values = c("T" = "coral", "RH" = "darkgoldenrod1"),
    labels = c("T" = "Temperature (°C)", "RH" = "Relative Humidity (g/m³)")
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
