---
title: "Stat107 Final Project"
subtitle: "How do harmful air pollutant concentrations vary across different months within a year?"
author: "Crystal Kuong, Gracelynne Mohad, Sumeet Sammi, Zoe Shum"
date: "2024-12-13"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-line-numbers: true
    page-layout: full
  pdf:
    fontsize: 10pt
    code-line-numbers: true
---

# Introduction

Air quality is a critical component in environmental health, directly influencing both the ecosystem and human well-being. It is determined by the presence of various pollutants including, but not limited to nitrogen dioxide and carbon monoxide. Poor air quality is linked to serious health issues, including respiratory diseases, cardiovascular problems, and reduced life expectancy. Moreover, air pollution contributes to climate change and harming wildlife. Monitoring and understanding air quality prompts initiatives centered on reducing the concentration of air pollutants, thereby, creating a cleaner environment.

## The Dataset

The UCI Air Quality Dataset was collected by the University of California, Irvine and acquired from Kaggle. It measures the concentration of about 5 different pollutants in the air in a given area. Some pollutants that are measured include carbon monoxide, benzene, and nitrogen dioxide. It measures pollutant concentrations every hour from March 2004 to April 2005.

### Packages

```{r}
library(readxl)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(dplyr)
```

### Reading Dataset

```{r}
air <- read_excel("AirQualityUCI.xls")
head(air)

# Number of rows and columns
dim(air)

# Column names
colnames(air)
```

The original dataset contains 9,357 observations and 15 variables.

1.  `Date`: The date of the measurement

2.  `Time`: The time of the measurement.

3.  `CO(GT)`: Concentration of carbon monoxide (CO) in the air (µg/m³).

4.  `PT0.S1(CO)`: Sensor measurement for CO concentration.

5.  `NMHC(GT)`: Concentration of non-methane hydrocarbons (NMHC) (µg/m³).

6.  `C6H6(GT)`: Concentration of benzene (C6H6) in the air (µg/m³).

7.  `PT08.S2(NMHC)`: Sensor measurement for NMHC concentration.

8.  `NOx(GT)`: Concentration of nitrogen oxide (NOx) in the air (µg/m³).

9.  `PT08.S3(NOx)`: Sensor measurement for NOx concentration.

10. `NO2(GT)`: Concentration of nitrogen dioxide (NO2) in the air (µg/m³).

11. `PT08.S4(NO2)`: Sensor measurement for NO2 concentration.

12. `PT08.S5(O3)`: Sensor measurement for ozone (O3) concentration

13. `T`: Temperature (Celsius)

14. `RH`: Relative humidity (g/m³)

15. `AH`: Absolute humidity (g/m³)

### Renaming Columns

Variable columns 3-12 will be renamed as followed to enhance clarity:

`CO(GT)` to `CO`

`PT0.S1(CO)` to `S_CO`

`NMHC(GT)` to `NMHC`

`PT08.S2(NMHC)` to `S_NMHC`

`C6H6(GT)` to `C6H6`

`NOx(GT)` to `NOx`

`PT08.S3(NOx)` to `S_NOx`

`NO2(GT)` to `NO2`

`PT08.S4(NO2)` to `S_NO2`

`PT08.S5(O3)` to `S_O3`

```{r}
air <- air %>% 
  rename(
    CO = `CO(GT)`,
    S_CO = `PT08.S1(CO)`,
    NMHC = `NMHC(GT)`,
    S_NMHC = `PT08.S2(NMHC)`,
    C6H6 = `C6H6(GT)`,
    NOx = `NOx(GT)`,
    S_NOx = `PT08.S3(NOx)`,
    NO2 = `NO2(GT)`,
    S_NO2 = `PT08.S4(NO2)`,
    S_O3 = `PT08.S5(O3)`
  )
```

### Tidying The Dataset

```{r}
#Summarize dataset (excluding Date and Time columns)
summary(air[, -c(1, 2)])
```

Based on the summary, it will be more helpful to implement the sensor measurements in the analyses since non-sensor measurements have negative means (which is unusual).

```{r}
# Histogram for S_CO 
hist(air$S_CO,
     main = "Histogram of CO concentrations",
     xlab = "CO concentrations")
```

From graphing a histogram of, it is obvious that outlines exist. We will use variable S_CO as a starting point for tidying up the data. We will use the IQR method to identify outliers and filter them out of the dataset.

```{r}
# Calculate IQR
Q1 <- quantile(air$S_CO, 0.25, na.rm = TRUE)
Q3 <- quantile(air$S_CO, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

# Define outlier bounds
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter rows to remove outliers
air_filtered <- air[air$S_CO >= lower_bound & air$S_CO <= upper_bound, ]
nrow(air)
nrow(air_filtered)
nrow(air) - nrow(air_filtered)
```

484 observations were removed from the dataset.

```{r}
hist(air_filtered$S_CO,
     main = "Histogram of CO concentrations (after filtering)",
     xlab = "CO concentrations")
```

while still positively skewed, it is a great improvement from the previous graph

```{r}
# Reassign original dataset with filtered dataset
air <- air_filtered
```

### Reformatting The Dataset

```{r}
# Convert 'Date' column to y-m-d format
air$Date <- ymd(air$Date)

# Create numerical Month and Year columns
air <- air %>%
  mutate(
    Month = month(Date, label = TRUE), # Extracts month as a numeric value (1-12)
    Year = year(Date)    # Extracts year
  )

# Group by Year and Month for summarization
airM <- air %>%
  group_by(Year, Month) %>%
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE)))

head(airM, 5)
```

# Methodology

# Analysis

```{r fig.width=10, fig.height=8}
# Plot monthly Concentrations
airM %>%
  # Use interaction to combine Month and Year
  ggplot(aes(x = interaction(Month, Year, sep = "-"))) +
  # Add points and lines for different pollutants
  geom_point(aes(y = S_CO, color = "Carbon Monoxide")) + 
  geom_line(aes(y = S_CO, color = "Carbon Monoxide", group = 1)) +
  geom_point(aes(y = S_NMHC, color = "Non-Methane Hydrocarbon")) +
  geom_line(aes(y = S_NMHC, color = "Non-Methane Hydrocarbon", group = 2)) +
  geom_point(aes(y = S_NOx, color = "Nitrogen Oxide")) + 
  geom_line(aes(y = S_NOx, color = "Nitrogen Oxide", group = 3)) +
  geom_point(aes(y = S_NO2, color = "Nitrogen Dioxide")) + 
  geom_line(aes(y = S_NO2, color = "Nitrogen Dioxide", group = 3)) +
  geom_point(aes(y = S_O3, color = "Ozone")) + 
  geom_line(aes(y = S_O3, color = "Ozone", group = 3)) +
  labs(
    title = "Average Concentrations of Pollutants by Month (2004-2005)",
    x = "Month-Year",
    y = "Concentrations in µg/m³"
  ) +
  scale_color_manual(
    values = c("Carbon Monoxide" = "blue", 
               "Non-Methane Hydrocarbon" = "red", 
               "Nitrogen Oxide" = "chartreuse3",
               "Nitrogen Dioxide" = "purple",
               "Ozone" = "magenta"),
    name = "Pollutants"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

airM %>%
  pivot_longer(
    cols = c(RH, T), 
    names_to = "Variable", 
    values_to = "Value"
  ) %>%
  ggplot(aes(x = interaction(Month, Year, sep = "-"), y = Value, fill = Variable)) +
  geom_col(position = "dodge") +
  labs(
    title = "Temperature and Relative Humidity by Month",
    x = "Month-Year",
    y = "Values",
    fill = "Variable"
  ) +
  scale_fill_manual(
    values = c("T" = "coral", "RH" = "darkgoldenrod1"),
    labels = c("T" = "Temperature (°C)", "RH" = "Relative Humidity (g/m³)")
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
<<<<<<< HEAD
```

```{r}
=======
>>>>>>> main
```
