---
title: "Stat107 Final Project"
subtitle: "How do harmful air pollutant concentrations vary across different months within a year?"
author: "Crystal Kuong, Gracelynne Mohad, Sumeet Sammi, Zoe Shum"
date: "2024-12-13"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-line-numbers: true
    page-layout: full
  pdf:
    fontsize: 10pt
    code-line-numbers: true
---

# Introduction

Air quality is a critical component in environmental health, directly influencing both the ecosystem and human well-being. It is determined by the presence of various pollutants including, but not limited to nitrogen dioxide and carbon monoxide. Poor air quality is linked to serious health issues, including respiratory diseases, cardiovascular problems, and reduced life expectancy. Moreover, air pollution contributes to climate change and harming wildlife. Monitoring and understanding air quality prompts initiatives centered on reducing the concentration of air pollutants, thereby, creating a cleaner environment.

## The Dataset

The UCI Air Quality Dataset was collected by the University of California, Irvine and acquired from Kaggle. It measures the concentration of about 5 different pollutants in the air in a given area. Some pollutants that are measured include carbon monoxide, benzene, and nitrogen dioxide. It measures pollutant concentrations every hour from March 2004 to April 2005.

```{r}
#Load libraries
library(readxl)
library(dplyr)
library(lubridate)
```

```{r}
#1: options(width = 50)
air <- read_excel("AirQualityUCI.xls")
head(air)

# Number of rows and columns
dim(air)

# Column names
colnames(air)
```

## Plan for Analysis

```{r}
# Convert 'Date' column to date format
air$Date <- ymd(air$Date) # "Parse dates with *y*ear, *m*onth, and *d*ay components"
air$Date <- format(air$Date, "%Y-%m") # Formats date to YYYY-mm 

# Group by month, where the average is calculated for each numeric column across all month groups, while removing all NA values
# summarise(): creates a new condensed dataset 
airM <- air %>%
  group_by(Date) %>%
  summarise(across(where(is.numeric), ~mean(., na.rm = TRUE))
)

head(airM, 5)
```
