---
title: "Stat107 Final Project"
subtitle: "How do harmful air pollutant concentrations vary across different months within a year?"
author: "Crystal Kuong, Gracelynne Mohan, Sumeet Sammi, Zoe Shum"
date: "2024-12-13"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
    code-line-numbers: true
    page-layout: full
  pdf:
    fontsize: 10pt
    code-line-numbers: true
---

# Introduction

Air quality is a vital aspect of environmental health, impacting ecosystems and human well-being. It is determined by the presence of various pollutants including, but not limited to nitrogen dioxide and carbon monoxide. Poor air quality is commonly linked to climate change and wildlife damage, as well as serious health issues, including respiratory diseases, cardiovascular problems, and reduced life expectancy. Understanding air quality trends can evoke initiatives centered on reducing the concentration of air pollutants, creating a cleaner environment and improving quality of life. 

## The Question

***This study investigates: How do harmful air pollutant concentrations vary across different months of the year?***

This study examines monthly variations in key air pollutants to identify patterns influenced by seasonal changes, human activities, or environmental factors. The pollutants analyzed include:

The specific pollutants we will explore in this study include:

1.  **Carbon Monoxide** ($CO$): A colorless, odorless gas from incomplete fossil fuel combustion, harmful to humans by impairing oxygen delivery.  

2.  **Non-Methane Hydrocarbon** ($NMHC$)**:** Organic compounds (excluding methane) that contribute to ground-level ozone and secondary pollutants through photochemical reactions. 

3.  **Nitrogen Dioxide** ($NO_2$)**:** A reddish-brown gas from fossil fuel combustion, linked to acid rain, smog, and respiratory issues.

4.  **Nitrogen Oxide** ($NOx$)**:** Gases from fossil fuel combustion that contribute to acid rain, smog, and ozone formation.

5.  **Ozone** ($O_3$)**:** A secondary pollutant formed from precursor emissions like $NOx$ and volatile organic compounds.

## Packages

```{r}
library(readxl)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(dplyr)
```

## The Dataset

The **UCI Air Quality Dataset**, sourced from the University of California, Irvine via Kaggle, tracks hourly pollutant, temperature, and humidity levels from March 2004 to April 2005 for a specific area.

```{r}
#| code-fold: true
air <- read_excel("AirQualityUCI.xls")
head(air)
```

The original dataset contained **9,357 observations** and **15 variables**. Variable columns 3-12 were renamed for clarity.

```{r}
#| code-fold: true
# Rename columns 3-12
air <- air %>% 
  rename(
    CO = `CO(GT)`,
    S_CO = `PT08.S1(CO)`,
    NMHC = `NMHC(GT)`,
    S_NMHC = `PT08.S2(NMHC)`,
    C6H6 = `C6H6(GT)`,
    NOx = `NOx(GT)`,
    S_NOx = `PT08.S3(NOx)`,
    NO2 = `NO2(GT)`,
    S_NO2 = `PT08.S4(NO2)`,
    S_O3 = `PT08.S5(O3)`
  )
```

| Variable | Description                                                    |
|------------------|------------------------------------------------------|
| `Date`   | The date of the measurement                                    |
| `Time`   | The time of the measurement.                                   |
| `CO`     | Concentration of carbon monoxide ($CO$) in the air (µg/m³).    |
| `S_CO`   | Sensor measurement for $CO$ concentration.                     |
| `NMHC`   | Concentration of non-methane hydrocarbons ($NMHC$) (µg/m³).    |
| `C6H6`   | Concentration of benzene ($C_6H_6$) in the air (µg/m³).        |
| `S_NMHC` | Sensor measurement for NMHC concentration.                     |
| `NOx`    | Concentration of nitrogen oxide ($NOx$) in the air (µg/m³).    |
| `S_NOx`  | Sensor measurement for $NOx$ concentration.                    |
| `NO2`    | Concentration of nitrogen dioxide ($NO_2$) in the air (µg/m³). |
| `S_NO2`  | Sensor measurement for $NO_2$ concentration.                   |
| `S_O3`   | Sensor measurement for ozone ($O_3$) concentration             |
| `T`      | Temperature (Celsius)                                          |
| `RH`     | Relative humidity (g/m³)                                       |
| `AH`     | Absolute humidity (g/m³)                                       |

### Tidying The Dataset

```{r}
#| code-fold: true
#Summarize dataset (excluding Date and Time columns)
summary(air[, -c(1, 2)])
```

```{r}
#| code-fold: true
# Histogram for S_CO 
hist(air$S_CO,
     main = "Histogram of CO concentrations",
     xlab = "CO concentrations")
```

The summary statistics revealed an unusual minimum value of -200.00 was observed across all variables.

To investigate this anomaly, the variable S_CO was analyzed as a test case, revealing the value to be an outlier. To address these issues and enhance data quality, outliers were identified and removed using the **Interquartile Range (IQR) method**.

```{r}
#| code-fold: true
# Calculate IQR
Q1 <- quantile(air$S_CO, 0.25, na.rm = TRUE)
Q3 <- quantile(air$S_CO, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

# Define outlier bounds
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter rows to remove outliers
air_filtered <- air[air$S_CO >= lower_bound & air$S_CO <= upper_bound, ]
```

The outcome:

-   **484** total observations were removed.

-   The distribution of `S_CO` shifted closer to normality, although still retaining a slight positive skew.

```{r}
#| code-fold: true
hist(air_filtered$S_CO,
     main = "Histogram of CO concentrations (after filtering)",
     xlab = "CO concentrations")
```

### Reformatting The Dataset

To effectively analyze monthly variations in pollutant concentrations, the dataset will be reorganized by grouping the data by month and year, calculating the average values for each variable within these groups. 

```{r}
#| code-fold: true
# Reassign original dataset with filtered dataset
air <- air_filtered

# Convert 'Date' column to y-m-d format
air$Date <- ymd(air$Date)

# Create numerical Month and Year columns
air <- air %>%
  mutate(
    Month = month(Date, label = TRUE), # Extracts month as a numeric value (1-12)
    Year = year(Date)    # Extracts year
  )

# Group by Year and Month for summarization
airM <- air %>%
  group_by(Year, Month) %>%
  summarize(across(where(is.numeric), 
                   ~mean(., na.rm = TRUE)),
            .groups = "drop")

head(airM, 5)
```

Upon inspecting the dataset, several non-sensor measurements revealed unusual negative values, suggesting potential inaccuracies. For this reason, the analysis will prioritize sensor-based variables.

```{r}
#| code-fold: true
# Disregard non-sensor pollutant measurements 
airM <- airM[ ,-c(3,5,6,8,10)]

head(airM, 5)
```

# Analysis

## Graphs

```{r fig.width=10, fig.height=6}
#| code-fold: true
# Plot monthly Concentrations
airM %>%
  # Use interaction to combine Month and Year
  ggplot(aes(x = interaction(Month, Year, sep = "-"))) +
  # Add points and lines for different pollutants
  geom_line(aes(y = S_CO, color = "Carbon Monoxide", group = 1)) +
  geom_line(aes(y = S_NMHC, color = "Non-Methane Hydrocarbon", group = 2)) +
  geom_line(aes(y = S_NOx, color = "Nitrogen Oxide", group = 3)) +
  geom_line(aes(y = S_NO2, color = "Nitrogen Dioxide", group = 3)) +
  geom_line(aes(y = S_O3, color = "Ozone", group = 3)) +
  labs(
    title = "Average Concentrations of Pollutants by Month (2004-2005)",
    x = "Month-Year",
    y = "Concentrations in µg/m³"
  ) +
  scale_color_manual(
    values = c("Carbon Monoxide" = "dodgerblue3", 
               "Non-Methane Hydrocarbon" = "brown2", 
               "Nitrogen Oxide" = "palegreen3",
               "Nitrogen Dioxide" = "mediumpurple3",
               "Ozone" = "hotpink2"),
    name = "Pollutant"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels
```

Nitrogen dioxide concentrations were notably higher than those of other pollutants from March to November 2004. While fluctuations were evident, certain trends stood out, such as a dip in August 2004 for ozone (pink), non-methane hydrocarbons (red), and carbon monoxide (blue). Additional dips occurred in February and April 2005, with the latter being more pronounced and severe. Interestingly, nitrogen oxide concentrations displayed a contrasting pattern compared to other pollutants. For example, in October 2004, when other pollutants peaked, nitrogen oxide concentrations reached their lowest levels.

```{r fig.width=10, fig.height=6}
#| code-fold: true
# Pivot data frame by Pollutant 
airM_2 <- airM %>%
  pivot_longer(
    cols = c(3:7), 
    names_to = "Pollutant", 
    values_to = "Concentration"
  )

# Plot Pollutant Concentrations 
airM_2 %>% 
  ggplot(aes(x = Pollutant)) +
  geom_boxplot(aes(y = Concentration, fill = Pollutant), size = 0.5) +
  labs(
    title = "Concentrations of Pollutants",
    x = "Pollutant",
    y = "Concentration (µg/m³)"
  ) +
  scale_fill_manual(
    values = c("dodgerblue3", "brown2", "mediumpurple3", "palegreen3", "hotpink2"), # Specify colors for each pollutant
    labels = c("Carbon Monoxide", "Non-methane Hydrocarbons", "Nitrogen Dioxide", "Nitrogen Oxide", "Ozone")
  )
```

The boxplot depicts the fluctuation of each pollutant. Most of the pollutants have a concentration lower than 1200 µg/m³. On the other hand, nitrogen dioxide had the highest concentration at above 1500 µg/m³. The high concentration of nitrogen dioxide is expected in urban areas where there is an abundance of fossil fuel combustion from cars. 

```{r fig.width=10, fig.height=6}
#| code-fold: true
ggplot() +
  geom_density(data = air, aes(x = S_CO, color="Carbon Monoxide")) +
  geom_density(data = air, aes(x = S_NMHC, color="Non-Methane Hydrocarbon")) +
  geom_density(data = air, aes(x = S_NOx, color="Nitrogen Oxide")) +
  geom_density(data = air, aes(x = S_NO2, color="Nitrogen Dioxide")) +
  geom_density(data = air, aes(x = S_O3, color="Ozone")) +
  labs(title = "Densities of Pollutant Concentrations", x = "Concentration (µg/m³)", y = "Density") +
  scale_color_manual(values = c("Carbon Monoxide" = "dodgerblue3", "Non-Methane Hydrocarbon" = "brown2", "Nitrogen Oxide" = "palegreen3","Nitrogen Dioxide" = "mediumpurple3","Ozone"  = "hotpink2"))


                 
```

The Densities of Pollution Concentrations graph shows the distribution of pollution concentrations in µg/m³. Carbon Monoxide and Nitrogen Oxide have the highest peaks, indicating that there’s a lot of measurements recorded in those specific concentration ranges.

Some of the more spread out distributions, such as Ozone and Nitrogen Dioxide, show that there’s a wider variation in the levels of concentration. Non-methane Hydrocarbons and Nitrogen Oxide seem to have a correlation, which is likely because they tend to coexist in areas with high pollution.

## Regression Analysis

```{r}
#| code-fold: true
# Generate the model first
model1 <- lm(log(S_NOx) ~ (S_CO), data = air)
summary(model1)

# Plot the relationship between CO and NO
air %>% 
  ggplot() +
  geom_point(aes(x = S_CO, y = S_NOx)) +  
  geom_line(
    aes(x = S_CO, y = exp(predict(model1, newdata = air))),  
    color = "green"
  ) +
  labs(
    title = "Relationship between CO and NO",
    x = "Carbon Monoxide",
    y = "Nitrogen Oxide"
  )

```

In an attempt to track the inverse relationship that was previously observed in the line graph between carbon monoxide and nitrogen oxide, a regression analysis was conducted. After plotting the data points, there seemed to be an exponential relationship between the two variables and a regression analysis produced a $R^2$ value of 0.69.

Therefore, there seems to be a moderate correlation between the variables. When carbon monoxide concentration levels go up, nitrogen oxide levels may be predicted to decrease at a decreasing rate, following a slow exponential decay.

# Conclusion

In conclusion, this study provided valuable insights into the variations in pollutant concentrations across different months and the seasonal trends influencing air quality. Notably, nitrogen dioxide levels were consistently high, likely due to the combustion of fossil fuels, while pollutants like ozone and carbon monoxide exhibited seasonal dips. Furthermore, the overall variation in pollutant levels across the year highlighted the dynamic nature of air quality, making identifying clear and consistent seasonal trends a challenge with the data currently available.

## Addressing Potential Errors

**Negative Values in the Dataset:** Negative values, such as “-200.00,” were assigned for missing or invalid data, leading to unrealistic negative means. To address this, we analyzed sensor measurements rather than actual pollutant concentrations. 

**Outliers in the Dataset:** Outliers were identified using the IQR method, but this approach can exclude valid extreme values, highlighting the need for careful consideration when filtering data. 

**Sensor Measurement Error:** Sensor errors due to calibration issues, environmental impacts, or sensor aging may result in inaccuracies, particularly in long-term data collection. 

**Regional Specificity:** This dataset focuses on the pollutant concentrations in a specific area, meaning that the variation of air quality across different locations may not be represented well. Factors such as traffic, industry, and geographical features can influence the pollutant levels in the area. While the data provides valuable insights into the air quality of the area over time, the data generalizations are limited as a result of the limited sampling region.
